---
title: "Heteroskedastic Gaussian Process Regressions"
author: "Mauna Dasari"
output:
  html_document:
    number_sections: no
    theme: paper
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F,warning=F,message=F,cache=2)
library(feather);library(tidyverse);library(kableExtra);library(gridExtra);

library(lme4);library(lmerTest);library(mgcv);library(survival);library(survminer);
metadata_all<-readRDS("../data/metadata_full.rds") %>% 
  rename(age_firstlivebirth=age_first_live_birth,
         age_firstpreg=age_first_preg,
         age_chrono=age.years) %>% 
  mutate(age_stat=(statdate-birth)/365.25)
```

# Intro

Given the heteroskedasticity in the data, we've decided that multiplying the variance by the gaussian process regression's kernel (rbf) is the most practical way to deal with it. This file uses a line to fit the relationship between predicted age and chronological age since it's easier to interpret. 


# Evaluating the Model {.tabset .tabset-pills}
## Plots {.tabset}
### All Animals
Pearson's R: 
```{r}
PredTest1<-read_feather("../data/H.All_PredictedAgeData_Test1.feather") #corr age.years and mb = 0.6913309
PredTest2<-read_feather("../data/H.All_PredictedAgeData_Test2.feather") #corr age.years and mb = 0.7007465
PredTest3<-read_feather("../data/H.All_PredictedAgeData_Test3.feather") #corr age.years and mb = 0.7024963
PredTest4<-read_feather("../data/H.All_PredictedAgeData_Test4.feather") #corr age.years and mb = 0.7029281
PredTest5<-read_feather("../data/H.All_PredictedAgeData_Test5.feather") #corr age.years and mb = 0.6939718

H.PredLife<-rbind(PredTest1,PredTest2,PredTest3,PredTest4,PredTest5) %>% 
  rename(age_chrono=age.years)
H.PredLife$delta<-H.PredLife$age_mb-H.PredLife$age_chrono
H.PredLife$delta_abs<-abs(H.PredLife$age_mb-H.PredLife$age_chrono)
saveRDS(H.PredLife,"../data/HetGPR.PredictedAgeData.rds")
cor(x=H.PredLife$age_chrono,y=H.PredLife$age_mb,method="pearson") #0.6983195
p.PredAge_Lifespan<-ggplot(data=H.PredLife, aes(x=age_chrono,y=age_mb,color=sex))+
  theme_minimal()+
  geom_abline(intercept = 0, slope = 1, linetype=2,colour="grey")+
  geom_point(pch=21,alpha=0.5)+
     geom_smooth(method=lm, fill=NA,fullrange = F,size=1.5)+
  scale_color_manual(name = 'sex',values = c('M' = '#115163','F' = 'darkgoldenrod3'),labels = c('female', 'male'))+
  xlim(0,30)+ylim(0,30)+
  xlab(bquote(Chronological~Age~(age[c])))+
  ylab(bquote(Microbial~Age~(age[m])))+
  theme(plot.title = element_text(hjust=0.5,size=16),
        aspect.ratio=1,
        legend.position = c(.2,.85), 
        legend.title=element_blank());p.PredAge_Lifespan

#ggsave(plot=p.PredAge_Lifespan,filename="../out/PredAgeLinear_sex_yellowblue.tiff",dpi=200)

fit<-lm(age_mb~age_chrono,data=H.PredLife)
H.PredLife$resid<-resid(fit)
fit2<-summary(fit)
RMSE=sqrt(mean(H.PredLife$age_chrono-H.PredLife$resid_lm)^2)
Adj.R_sq=fit2$r.sq
Variance=(1-fit2$r.sq)
PearsonsR=cor(x=H.PredLife$age_chrono,y=H.PredLife$age_mb,method="pearson")
MedianErr=median(H.PredLife$delta_abs)
SD=sd(H.PredLife$age_chrono)
AvgAcceleration=mean(H.PredLife$delta)
CV_chrono=raster::cv(H.PredLife$age_chrono)
CV_mb=raster::cv(H.PredLife$age_mb)
Aggregate8020_All<-c("13476","13476",RMSE,Adj.R_sq,Variance,SD,CV_chrono,CV_mb,PearsonsR,MedianErr,AvgAcceleration)
```

### Female Subset
Pearson's R: 
```{r}
H.PredLife_F<-H.PredLife%>% filter(sex=="F")

cor(x=H.PredLife_F$age_chrono,y=H.PredLife_F$age_mb,method="pearson")
p.PredAge_Lifespan<-ggplot(data=H.PredLife_F, aes(x=age_chrono,y=age_mb,color=sex))+
  theme_minimal()+
  geom_abline(intercept = 0, slope = 1, linetype=2,colour="grey")+
  geom_point(pch=21,alpha=0.5)+
     geom_smooth(method=lm, fill=NA,fullrange = F,size=1.5)+
  scale_color_manual(name = 'sex',values = c('M' = '#115163','F' = 'darkgoldenrod3'),labels = c('female', 'male'))+
  xlim(0,30)+ylim(0,30)+
  xlab(bquote(Chronological~Age~(age[c])))+
  ylab(bquote(Microbial~Age~(age[m])))+
  labs(title="GPR (RBF*var(y)), Linear Fit")+
  theme(plot.title = element_text(hjust=0.5,size=16),
        aspect.ratio=1,
        legend.position = c(.2,.85), 
        legend.title=element_blank());p.PredAge_Lifespan

fit<-lm(age_mb~age_chrono,data=H.PredLife_F)
H.PredLife_F$resid<-resid(fit)
fit2<-summary(fit)
RMSE=sqrt(mean(H.PredLife_F$age_chrono-H.PredLife_F$resid_lm)^2)
Adj.R_sq=fit2$r.sq
Variance=(1-fit2$r.sq)
PearsonsR=cor(x=H.PredLife_F$age_chrono,y=H.PredLife_F$age_mb,method="pearson")
MedianErr=median(H.PredLife_F$delta_abs)
SD=sd(H.PredLife_F$age_chrono)
AvgAcceleration=mean(H.PredLife_F$delta)
CV_chrono=raster::cv(H.PredLife_F$age_chrono)
CV_mb=raster::cv(H.PredLife_F$age_mb)
Aggregate8020_F<-c("8245","8245",RMSE,Adj.R_sq,Variance,SD,CV_chrono,CV_mb,PearsonsR,MedianErr,AvgAcceleration)
```

### Male Subset
Pearson's R: 
```{r}
H.PredLife_M<-H.PredLife %>% filter(sex=="M")

cor(x=H.PredLife_M$age_chrono,y=H.PredLife_M$age_mb,method="pearson")
p.PredAge_Lifespan<-ggplot(data=H.PredLife_M, aes(x=age_chrono,y=age_mb,color=sex))+
  theme_minimal()+
  geom_abline(intercept = 0, slope = 1, linetype=2,colour="grey")+
  geom_point(pch=21,alpha=0.5)+
     geom_smooth(method=lm, fill=NA,fullrange = F,size=1.5)+
  scale_color_manual(name = 'sex',values = c('M' = '#115163','F' = 'darkgoldenrod3'),labels = c('male', 'female'))+
  xlim(0,30)+ylim(0,30)+
  xlab(bquote(Chronological~Age~(age[c])))+
  ylab(bquote(Microbial~Age~(age[m])))+
  labs(title="GPR (RBF*var(y)), Linear Fit")+
  theme(plot.title = element_text(hjust=0.5,size=16),
        aspect.ratio=1,
        legend.position = c(.2,.85), 
        legend.title=element_blank());p.PredAge_Lifespan

fit<-lm(age_mb~age_chrono,data=H.PredLife_M)
H.PredLife_M$resid<-resid(fit)
fit2<-summary(fit)
RMSE=sqrt(mean(H.PredLife_M$age_chrono-H.PredLife_M$resid_lm)^2)
Adj.R_sq=fit2$r.sq
Variance=(1-fit2$r.sq)
PearsonsR=cor(x=H.PredLife_M$age_chrono,y=H.PredLife_M$age_mb,method="pearson")
MedianErr=median(H.PredLife_M$delta_abs)
SD=sd(H.PredLife_M$age_chrono)
AvgAcceleration=mean(H.PredLife_M$delta)
CV_chrono=raster::cv(H.PredLife_M$age_chrono)
CV_mb=raster::cv(H.PredLife_M$age_mb)
Aggregate8020_M<-c("5231","5231",RMSE,Adj.R_sq,Variance,SD,CV_chrono,CV_mb,PearsonsR,MedianErr,AvgAcceleration)
```

## Summary Statistics
```{r}
Sum8020<-as.data.frame(rbind(Aggregate8020_All,Aggregate8020_F,Aggregate8020_M))
colnames(Sum8020)<-c("n_training","n_test","RMSE","Adj.R_sq","Variance","SD","CV_chrono","CV_mb","PearsonsR","MedianErr","AvgAcceleration")
Sum8020<-Sum8020 %>% 
  mutate_if(is.factor,as.character) %>% 
  mutate_if(is.character, as.numeric) %>% 
  mutate(Model=c("All Animals","Female Subset","Male Subset"))
Sum8020<-Sum8020[,c(12,1:11)]
#nice green #517C63
Sum8020%>% 
  kable(digits = 3,format = 'html', escape = F) %>%
  kable_styling() %>% 
  kable_styling(fixed_thead = T) 

# Sex interaction 
summary(lm(age_mb ~ age_chrono + sex + age_chrono:sex,data=H.PredLife))

# Which sex ages faster? 
summary(lm(age_mb ~ age_chrono,data=H.PredLife[H.PredLife$sex=="F",]))

summary(lm(age_mb ~ age_chrono,data=H.PredLife[H.PredLife$sex=="M",]))
```

