---
title: "ENR For Release"
author: "Mauna Dasari"
date: "1/4/2020"
output: 
  html_document:
    theme: spacelab
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F,warning=F,message=F,cache=2)

library(glmnet);library(tidyverse);library(survival);library(survminer); library(dplyr);library(rlang);library(ggplot2);library(lmerTest);library(gridExtra); library(kableExtra)

#phenotypes<-readRDS('../out/phenotypes_CLRscaled_no_chl_mt.rds')
metadata_all<-readRDS("../out/metadata_bstatus0_mathormonerank_rainpriortomilestone_matrankbirthgroup.rds") %>% 
  rename(age_firstlivebirth=age_first_live_birth,
         age_firstpreg=age_first_preg)
metadata_random<-readRDS("../out/metadata_random.rds")

theme2<-theme(plot.background = element_blank(), 
  panel.grid.major = element_blank(), 
  panel.grid.minor = element_blank(), 
  #panel.border = element_blank(), 
  panel.background = element_blank(),
  axis.line.x = element_line(size=.4), 
  axis.line.y = element_line(size=.4), 
  legend.position="none",
  axis.title.x = element_text(color="black", size=10),
  axis.title.y = element_text(color="black", size=10),
  plot.title=element_text(hjust=0.5, size=12))
```

# Data Generation {.tabset}

Here is the code for the finalized elastic net regressions. After some optimizing for the 80:20 model, I settled on the following parameters:
* alpha = 0.2 (closer to a ridge regression)
* nfolds = 75
* data from entire lifespan

## ENRs {.tabset}
### Nested Cross Validation 80:20 with Samples {.tabset .tabset-pills}

#### Make the data
```{r eval=F, echo=T}
metadata_random<-readRDS("../out/metadata_random.rds")

#%%%%%%%%%%%%%%%#
#  INDIVIDUALS  #
#%%%%%%%%%%%%%%%#
n_test=5 #5 for 80:20 split
#Filter out those with only one sample in the time frame.
##Figure out who those individuals are
metadata_filtered_no_singletons_lifespan<-metadata_random %>% 
  group_by(sname,sex,fake_sex) %>% 
  summarize(count=n_distinct(DADA_id)) %>% #479 individuals
  ungroup() %>% 
  filter(count>1) %>%  #431 individuals
  mutate(testID=sample(1:n_test,size=431,replace=T)) %>%  #hard coded 431 so I wouldn't have to create a new table
  ungroup()
##Filter out individuals not on the list
info<-metadata_random %>% 
  select("DADA_id","sname","age.years","sex","fake_sex") %>%
  filter(sname %in% metadata_filtered_no_singletons_lifespan$sname) %>%
  left_join(metadata_filtered_no_singletons_lifespan, by=c("sname","sex","fake_sex"))
saveRDS(info,"../out/ENR_Final/metadata_filtered_8020individuals.rds")

#%%%%%%%%%%%#
#  SAMPLES  #
#%%%%%%%%%%%#

#Filter out those with only one sample in the time frame.
##Figure out who those individuals are
metadata_filtered_no_singletons_lifespan<-metadata_random %>% 
  group_by(sname,sex,fake_sex) %>% 
  summarize(count=n_distinct(DADA_id)) %>% #479 individuals
  ungroup() %>% 
  filter(count>1) #431 individuals
##Filter out individuals not on the list
n_test=5 #5 for 80:20 split
info<-metadata_random %>% 
  select("DADA_id","sname","age.years","sex","fake_sex") %>%
  filter(sname %in% metadata_filtered_no_singletons_lifespan$sname) %>%
  group_by(sname) %>% 
  mutate(count=n(),
         testID=ifelse(count>n_test,
                       sample(1:n_test,size=count,replace=T),
                       sample(1:n_test,size=count,replace=F))) %>% 
  ungroup()
saveRDS(info,"../out/ENR_Final/metadata_filtered_8020.rds")
```

```{r}
info<-readRDS("../out/ENR_Final/metadata_filtered_8020individuals.rds") %>% 
  group_by(sname,testID) %>% 
  summarize() %>% 
  ungroup()

hist(info$testID)
```

#### ENR

This ENR is written so I can easily switch between a randomization at the sample level or the individual level.
```{r eval=F, echo=T}
library(glmnet)
library(tibble)
library(dplyr)
library(rlang)
library(doParallel)
registerDoParallel(16)

phenotypes<-readRDS("../../raw_data/phenotypes_CLRscaled_no_chl_mt.rds")
#info<-readRDS("../../raw_data/metadata_filtered_8020.rds");n=5;folder="NestedCV8020";
info<-readRDS("../../raw_data/metadata_filtered_8020individuals.rds");n=5;folder="NestedCV8020Individuals";
# creating the extra tables we need #
SummaryStats_list<-c("sname","n_Training","n_Test","Intercept",
                          "MSE","RMSE","RSE","R_sq","Adj.R_sq",
                          "R_sq_1to1","Variance","t_ratio","SD", 
                          "PearsonsR","MedianErr","AvgAcceleration")
SummaryStats<-data.frame(matrix(ncol=length(SummaryStats_list),nrow=n))
colnames(SummaryStats)<-SummaryStats_list

#%%%%%%%%%%%%%#
# ALL ANIMALS #
#%%%%%%%%%%%%%#

for(i in 1:max(info$testID)){

#Filter the test and training data
train_info<-info %>% 
  filter(testID!=i) %>% 
  arrange(.[['DADA_id']])

test_info<-info %>% 
  filter(testID==i) %>% 
  arrange(.[['DADA_id']])

#Create features tables for both training and test data
train_count<-phenotypes %>% 
  rownames_to_column("DADA_id") %>% 
  as.tibble() %>% 
  filter(.[['DADA_id']] %in% train_info$DADA_id) %>% 
  arrange(.[['DADA_id']]) %>% 
  column_to_rownames("DADA_id")

test_count<-phenotypes %>% 
  rownames_to_column("DADA_id") %>% 
  as.tibble() %>% 
  filter(.[['DADA_id']] %in% test_info$DADA_id) %>% 
  arrange(.[['DADA_id']]) %>% 
  column_to_rownames("DADA_id")

#Run model
set.seed(42)
model<-cv.glmnet(as.matrix(train_count),
                 train_info$age.years,
                 nfolds=75,
                 alpha=.20,
                 standardize=F,
                 parallel=T)

#Use model to predict test data
PredictedAge<-predict(model,
                       newx=as.matrix(test_count),
                       s=model$lambda.min)

#Pull weights
weights<-as.data.frame(coef(model,s=model$lambda.min)[,1])
rownames(weights)<-c("Intercept",colnames(phenotypes))
name_weights<-paste("../../out/ENR_Final/", folder, "/Weights75_Test",i,".rds",sep = "")
saveRDS(weights,name_weights)

#Pull test predictions
test_info_postmodel<-as.data.frame(cbind(test_info$age.years,PredictedAge))
colnames(test_info_postmodel)<-c("age_chrono","age_mb")
test_info_postmodel$delta<-(test_info_postmodel$age_chrono-test_info_postmodel$age_mb)
test_info_postmodel$delta_abs<-abs(test_info_postmodel$age_chrono-test_info_postmodel$age_mb)
test_info_postmodel$DADA_id<-rownames(test_info_postmodel)

PredictedAgeData<-data.frame(matrix(ncol=5,nrow=1))
colnames(PredictedAgeData)<-c("age_chrono","age_mb","delta","delta_abs","DADA_id")
PredictedAgeData<-rbind(PredictedAgeData,test_info_postmodel)
PredictedAgeData = PredictedAgeData[-1,]
PredictedAgeData<-PredictedAgeData[,c(5,1:4)]
name_predict<-paste("../../out/ENR_Final/", folder, "/PredictedAgeData75_Test",i,".rds", sep="")
saveRDS(PredictedAgeData,name_predict)

#Take summary metrics
fit<-summary(lm(PredictedAge~test_info$age.years))
SummaryStats[i,1]="All Lifespan"
SummaryStats[i,2]=(length(train_info$age.years))
SummaryStats[i,3]=(length(test_info$age.years))
SummaryStats[i,4]=fit$coefficients[1,1]
SummaryStats[i,5]=mean((test_info$age.years-residuals(fit))^2)
SummaryStats[i,6]=(sqrt(mean(test_info$age.years-residuals(fit))^2))
SummaryStats[i,7]=fit$sigma
SummaryStats[i,8]=fit$r.squared
SummaryStats[i,9]=fit$adj.r.squared
SummaryStats[i,10]=1 - (with( test_info_postmodel,sum( age_mb-age_chrono)^2 ))/with(test_info_postmodel, sum( age_mb^2))
SummaryStats[i,11]=(1-fit$r.squared)
SummaryStats[i,12]=(fit$coefficients[2,1]-1)/fit$coefficients[2,2]
SummaryStats[i,13]=sd(test_info_postmodel$age_chrono)
SummaryStats[i,14]=cor(x=test_info_postmodel$age_chrono,y=test_info_postmodel$age_mb,
                            method="pearson")
SummaryStats[i,15]=median(test_info_postmodel$delta_abs)
SummaryStats[i,16]=mean(test_info_postmodel$delta)
}
name_summary<-paste("../../out/ENR_Final/", folder, "/SummaryStats75.rds", sep="")
saveRDS(SummaryStats,name_summary)
```

# Comparing the types of nested cross validations {.tabset .tabset-pills}
This section shows the plots I used to compare the types of nested cross-validation (80:20 at the sample level, 80:20 at the individual level, and focal individual as test)

## 80:20 Sample Nested CV {.tabset}
Number above the graphs corresponds to Pearson's R between the predicted age and chronological age.
```{r}
PredTest1<-readRDS("../out/ENR_Final/NestedCV8020Samples/Lifespan/PredictedAgeData75_Test1.rds");PredTest1$Test_Set<-"Test1"
PredTest2<-readRDS("../out/ENR_Final/NestedCV8020Samples/Lifespan/PredictedAgeData75_Test2.rds");PredTest2$Test_Set<-"Test2"
PredTest3<-readRDS("../out/ENR_Final/NestedCV8020Samples/Lifespan/PredictedAgeData75_Test3.rds");PredTest3$Test_Set<-"Test3"
PredTest4<-readRDS("../out/ENR_Final/NestedCV8020Samples/Lifespan/PredictedAgeData75_Test4.rds");PredTest4$Test_Set<-"Test4"
PredTest5<-readRDS("../out/ENR_Final/NestedCV8020Samples/Lifespan/PredictedAgeData75_Test5.rds");PredTest5$Test_Set<-"Test5"

metadata_min<-metadata_random %>% select("DADA_id","sname","sex","fake_sex")
PredLife<-rbind(PredTest1,PredTest2,PredTest3,PredTest4,PredTest5) %>% left_join(metadata_min,by="DADA_id")
#saveRDS(PredLife,"../out/ENR_Final/PredLife_8020SampleModelAggregate.rds")

cor(x=PredLife$age_chrono,y=PredLife$age_mb,method="pearson")
p.PredAge_Lifespan<-ggplot(data=PredLife, aes(x=age_chrono,y=age_mb,color=sex))+
  theme_minimal()+
  geom_abline(intercept = 0, slope = 1, linetype=2,colour="grey")+
  geom_point(pch=21,alpha=0.5)+
     geom_smooth(method=lm, fill=NA,fullrange = F,size=1.5)+
  scale_color_manual(name = 'sex',values = c('M' = '#115163','F' = 'darkgoldenrod3'),labels = c('female', 'male'))+
  #xlim(0,30)+ylim(0,30)+
  xlab(bquote(Chronological~Age~(age[c])))+
  ylab(bquote(Microbial~Age~(age[m])))+
  labs(title="All Lifespan (n=13,476), Linear Fit by Sex")+
  theme(plot.title = element_text(hjust=0.5,size=16),
        aspect.ratio=1,
        legend.position = c(.2,.85), 
        legend.title=element_blank());p.PredAge_Lifespan
#ggsave(plot=p.PredAge_Lifespan,filename="../out/plot/Sample8020_PredAge_linear.tiff",dpi=200)
p.PredAge_Lifespan2<-ggplot(data=PredLife, aes(x=age_chrono,y=age_mb,color=sex))+
  theme_minimal()+
  geom_abline(intercept = 0, slope = 1, linetype=2,colour="grey")+
  geom_point(pch=21,alpha=0.5)+
     geom_smooth(fill=NA,fullrange = F,size=1.5)+
  scale_color_manual(name = 'sex',values = c('M' = '#115163','F' = 'darkgoldenrod3'),labels = c('female', 'male'))+
  #xlim(0,30)+ylim(0,30)+
  xlab(bquote(Chronological~Age~(age[c])))+
  ylab(bquote(Microbial~Age~(age[m])))+
  labs(title="All Lifespan (n=13,476), GAM Fit by Sex")+
  theme(plot.title = element_text(hjust=0.5,size=16),
        aspect.ratio=1,
        legend.position = c(.2,.85), 
        legend.title=element_blank());p.PredAge_Lifespan2
#ggsave(plot=p.PredAge_Lifespan,filename="../out/plot/Sample8020_PredAgePlot_sex.tiff", 
#       dpi=200,width=7,height=7)
#ggsave(plot=p.PredAge_Lifespan2,filename="../out/plot/Sample8020_PredAge_gam.tiff",dpi=200)
p.PredAge_Lifespan.natural<-ggplot(data=PredLife, aes(x=age_chrono,y=age_mb,color=sex))+
  theme_minimal()+
  geom_abline(intercept = 0, slope = 1, linetype=2,colour="grey")+
  geom_point(pch=21,alpha=0.5)+
     geom_smooth(method=lm, fill=NA,fullrange = F,size=1.5)+
  scale_color_manual(name = 'sex',values = c('M' = '#115163','F' = 'darkgoldenrod3'),labels = c('female', 'male'))+
  labs(title="All Lifespan (n=13,476)")+
  xlab(bquote(Chronological~Age~(age[c])))+
  ylab(bquote(Microbial~Age~(age[m])))+
  theme(plot.title = element_text(hjust=0.5,size=16),
        aspect.ratio=1,
        legend.position = "none", 
        legend.title=element_blank());p.PredAge_Lifespan.natural

fit<-summary(lm(PredLife$age_mb~PredLife$age_chrono))
Intercept=fit$coefficients[1,1]
MSE=mean((PredLife$age_chrono-residuals(fit))^2)
RMSE=(sqrt(mean(PredLife$age_chrono-residuals(fit))^2))
RSE=fit$sigma
R_sq=fit$r.squared
Adj.R_sq=fit$adj.r.squared
R_sq_1to1=1 - (with( PredLife,sum( age_mb-age_chrono)^2 ))/with(PredLife, sum( age_mb^2))
Variance=(1-fit$r.squared)
t_ratio=(fit$coefficients[2,1]-1)/fit$coefficients[2,2]
SD=sd(PredLife$age_chrono)
PearsonsR=cor(x=PredLife$age_chrono,y=PredLife$age_mb,
                            method="pearson")
MedianErr=median(PredLife$delta_abs)
AvgAcceleration=mean(PredLife$delta)
Aggregate8020_Lifespan<-c("13456","13456",Intercept,MSE,RMSE,RSE,R_sq,Adj.R_sq,R_sq_1to1,Variance,t_ratio,SD,PearsonsR,MedianErr,AvgAcceleration)

m.lifespan<-summary(lm(data=PredLife,age_mb~age_chrono+sex+sex*age_chrono))
Lifespan.estimates<-m.lifespan$coefficients[,1];Lifespan.se<-m.lifespan$coefficients[,2];Lifespan.pvalue<-m.lifespan$coefficients[,4];
m.lifespan;Lifespan.estimates;Lifespan.se
```

## 80:20 Individual Nested CV {.tabset}
Number above the graphs corresponds to Pearson's R between the predicted age and chronological age.
```{r}
PredTest1<-readRDS("../out/ENR_Final/NestedCV8020Individuals/PredictedAgeData75_Test1.rds");PredTest1$Test_Set<-"Test1"
PredTest2<-readRDS("../out/ENR_Final/NestedCV8020Individuals/PredictedAgeData75_Test2.rds");PredTest2$Test_Set<-"Test2"
PredTest3<-readRDS("../out/ENR_Final/NestedCV8020Individuals/PredictedAgeData75_Test3.rds");PredTest3$Test_Set<-"Test3"
PredTest4<-readRDS("../out/ENR_Final/NestedCV8020Individuals/PredictedAgeData75_Test4.rds");PredTest4$Test_Set<-"Test4"
PredTest5<-readRDS("../out/ENR_Final/NestedCV8020Individuals/PredictedAgeData75_Test5.rds");PredTest5$Test_Set<-"Test5"

metadata_min<-metadata_random %>% select("DADA_id","sname","sex","fake_sex")
PredLife_Individuals<-rbind(PredTest1,PredTest2,PredTest3,PredTest4,PredTest5) %>% left_join(metadata_min,by="DADA_id")

cor(x=PredLife_Individuals$age_chrono,y=PredLife_Individuals$age_mb,method="pearson")
p.PredAge_Lifespan_Individuals<-ggplot(data=PredLife_Individuals, aes(x=age_chrono,y=age_mb,color=sex))+
  theme_minimal()+
  geom_abline(intercept = 0, slope = 1, linetype=2,colour="grey")+
  geom_point(pch=21,alpha=0.5)+
     geom_smooth(method=lm, fill=NA,fullrange = T)+
  scale_color_manual(name = 'sex',values = c('M' = '#115163','F' = 'darkgoldenrod3'),labels = c('female', 'male'))+
  xlim(0,30)+ylim(0,30)+
  xlab(bquote(Chronological~Age~(age[c])))+
  ylab(bquote(Microbial~Age~(age[m])))+
  labs(title="All Lifespan (n=13,476)")+
  theme(plot.title = element_text(hjust=0.5,size=12),
        aspect.ratio=1,
        legend.position = c(.2,.85), 
        legend.title=element_blank());p.PredAge_Lifespan_Individuals
#(plot=p.PredAge_Lifespan_Individuals,filename="../out/plot/Indiv8020_PredAgePlot_sex.tiff", 
#       dpi=200,width=7,height=7)

p.PredAge_Lifespan.natural_Individuals<-ggplot(data=PredLife_Individuals, aes(x=age_chrono,y=age_mb,color=sex))+
  theme_minimal()+
  geom_abline(intercept = 0, slope = 1, linetype=2,colour="grey")+
  geom_point(pch=21,alpha=0.5)+
     geom_smooth(method=lm, fill=NA,fullrange = T)+
  scale_color_manual(name = 'sex',values = c('M' = '#115163','F' = 'darkgoldenrod3'),labels = c('female', 'male'))+
  labs(title="All Lifespan (n=13,476)")+
  xlab(bquote(Chronological~Age~(age[c])))+
  ylab(bquote(Microbial~Age~(age[m])))+
  theme(plot.title = element_text(hjust=0.5,size=12),
        aspect.ratio=1,
        legend.position = "none", 
        legend.title=element_blank());p.PredAge_Lifespan.natural_Individuals

fit<-summary(lm(PredLife_Individuals$age_mb~PredLife_Individuals$age_chrono))
Intercept=fit$coefficients[1,1]
MSE=mean((PredLife_Individuals$age_chrono-residuals(fit))^2)
RMSE=(sqrt(mean(PredLife_Individuals$age_chrono-residuals(fit))^2))
RSE=fit$sigma
R_sq=fit$r.squared
Adj.R_sq=fit$adj.r.squared
R_sq_1to1=1 - (with( PredLife_Individuals,sum( age_mb-age_chrono)^2 ))/with(PredLife_Individuals, sum( age_mb^2))
Variance=(1-fit$r.squared)
t_ratio=(fit$coefficients[2,1]-1)/fit$coefficients[2,2]
SD=sd(PredLife_Individuals$age_chrono)
PearsonsR=cor(x=PredLife_Individuals$age_chrono,y=PredLife_Individuals$age_mb,
                            method="pearson")
MedianErr=median(PredLife_Individuals$delta_abs)
AvgAcceleration=mean(PredLife_Individuals$delta)
Aggregate8020_Lifespan_Individuals<-c("13456","13456",Intercept,MSE,RMSE,RSE,R_sq,Adj.R_sq,R_sq_1to1,Variance,t_ratio,SD,PearsonsR,MedianErr,AvgAcceleration)
```

