---
title: "Phenotype Testing - Comparing Linear and Quadratic Models Across Lifespan"
author: "Mauna Dasari"
output: 
  html_document:
    theme: spacelab
    toc: false
    toc_depth: 3
    toc_float: true
---
```{r setup, include=FALSE, eval=T}
knitr::opts_chunk$set(echo = F, warning=F,message=F,cache=2,
                      fig.width=20,fig.height=15)

library(tidyverse); library(dplyr); library(kableExtra);library(knitr);
library(vegan);library(grid);library(gridExtra);library(lme4);library(lmerTest);
library(beepr);library(pander);library(wesanderson); library(LaCroixColoR)

theme1<-theme_bw()+
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "none")

theme2<-theme(                              
  plot.background = element_blank(), 
  panel.grid.major = element_blank(), 
  panel.grid.minor = element_blank(), 
  panel.background = element_blank(),
  axis.line.x = element_line(size=.4), 
  axis.line.y = element_line(size=.4), 
  legend.position="none",
  axis.title.x = element_text(color="black", size=10),
  axis.title.y = element_text(color="black", size=10),
  plot.title=element_text(hjust=0.5, size=12))
```

---

```{r metadata_z_score, eval=T}
metadata_raw<-readRDS("../../1_data/out/metadata_with_pc_points_and_diet.rds")
metadata_all<-metadata_raw%>%
    mutate(read_count.z=((read_count-mean(read_count, na.rm=T))/sd(read_count,na.rm=T)))%>%
    #mutate(age.years.z=((age.years - mean(age.years, na.rm=T))/sd(age.years,na.rm=T)))%>%
    mutate(rain_1moprior.z=((rain_1moprior-mean(rain_1moprior,na.rm=T))/sd(rain_1moprior,na.rm=T)))%>%
    mutate(rain_3moprior.z=((rain_3moprior-mean(rain_3moprior,na.rm=T))/sd(rain_3moprior,na.rm=T)))%>%
    mutate(rain_1yrprior.z=((rain_1yrprior-mean(rain_1yrprior,na.rm=T))/sd(rain_1yrprior, na.rm=T)))%>%
    mutate(t_log=case_when(sex=="M"~as.numeric(t_log)))%>%
    mutate(e2_log=case_when(sex=="F"~as.numeric(e2_log)))%>% #getting ready to standardize within sex
    mutate(e2_log.z=(e2_log-mean(e2_log,na.rm=T))/sd(e2_log,na.rm=T))%>%
    mutate(t_log.z=(t_log-mean(t_log,na.rm=T))/sd(t_log,na.rm=T))%>%
    mutate(gc_log.z=(gc_log-mean(gc_log,na.rm=T))/sd(gc_log,na.rm=T))%>%
    mutate(diet_shannon_h.z=(diet_shannon_h-mean(diet_shannon_h,na.rm=T))/sd(diet_shannon_h,na.rm=T))%>%
    mutate(diet_PC1.z=(diet_PC1-mean(diet_PC1,na.rm=T))/sd(diet_PC1,na.rm=T))%>%
    mutate(diet_PC2.z=(diet_PC2-mean(diet_PC2,na.rm=T))/sd(diet_PC2,na.rm=T))%>%
    mutate(diet_PC3.z=(diet_PC3-mean(diet_PC3,na.rm=T))/sd(diet_PC3,na.rm=T))%>%
    mutate(diet_PC4.z=(diet_PC4-mean(diet_PC4,na.rm=T))/sd(diet_PC4,na.rm=T))%>%
    mutate(diet_PC5.z=(diet_PC5-mean(diet_PC5,na.rm=T))/sd(diet_PC5,na.rm=T))%>%
    mutate(e2_log.z=case_when(sex=="F"~as.numeric(e2_log.z), #setting NA = 0 to make sure model works
                              sex=="M"~0))%>%
    mutate(t_log.z=case_when(sex=="M"~as.numeric(t_log.z),
                             sex=="F"~0))%>%
    dplyr::select(-c(matgrp,rain_3moprior,
                     bstatus,matgrp,sid,did,
                     matgrp,day_of_collection,e2, gc,
                     t,e2_log,t_log,gc_log,e2_date, t_date,
                     gc_date,season,contains("diet_PC*")))

metadata<-metadata_all%>%
  mutate(e2_log.z=case_when(sex=="F"~as.numeric(e2_log.z), #setting NA = 0 to make sure model works
                              sex=="M"~0))%>%
  mutate(t_log.z=case_when(sex=="M"~as.numeric(t_log.z),
                              sex=="F"~0))%>%
  mutate(matstate=case_when(collection_date>=matured~'adult',
                            collection_date<matured~'juvenile',
                            TRUE~'juvenile'))
metadata$matstate<-factor(metadata$matstate,levels=c("juvenile","adult"))
#saveRDS(metadata,"../out/microbiome16S_metadata.rds")
```

```{r metadata_no_z,eval=F}
metadata_raw<-readRDS("../../1_data/out/metadata_with_pc_points_and_diet.rds")

metadata<-metadata_raw%>%
  mutate(e2_log=case_when(sex=="F"~as.numeric(e2_log), #setting NA = 0 to make sure model works
                              sex=="M"~0))%>%
  mutate(t_log=case_when(sex=="M"~as.numeric(t_log),
                              sex=="F"~0))%>%
  dplyr::select(-c(matgrp,collect_month,collect_year,grp_month,rain_3moprior,rain_6moprior,bstatus,matgrp,sid,did,matgrp,day_of_collection,e2,gc,t,e2_date, t_date, gc_date))%>%
  mutate(matstate=case_when(collection_date>=matured~'adult',
                            collection_date<matured~'juvenile',
                            TRUE~'juvenile'))
metadata$matstate<-factor(metadata$matstate,levels=c("juvenile","adult"))
```

```{r metadata_other}

#For Jordan
microbiome_sampleID_age<-metadata%>%
  select(DADA_id,sname,age.years,sex)
saveRDS(microbiome_sampleID_age, "../out/microbiome16S_sampleID_sname_age.rds")
microbiome_sampleID<-microbiome_sampleID_age%>%
  select(DADA_id)
phylum<-readRDS("../../1_data/out/reads_phylum.rds")
phylum$DADA_id<-rownames(phylum)
family<-readRDS("../../1_data/out/reads_family.rds")
family$DADA_id<-rownames(family)
genus<-readRDS("../../1_data/out/reads_genus.rds")
genus$DADA_id<-rownames(genus)
genus<-readRDS("../../1_data/out/relabundance_genus_scaled.rds")
genus$DADA_id<-rownames(genus)

microbiome_sampleID_taxa_abundances<-merge(microbiome_sampleID,phylum, by="DADA_id")
microbiome_sampleID_taxa_abundances<-merge(microbiome_sampleID_taxa_abundances,family, by="DADA_id")
microbiome_sampleID_taxa_abundances<-merge(microbiome_sampleID_taxa_abundances,genus, by="DADA_id")
saveRDS(microbiome_sampleID_taxa_abundances,"../out/microbiome16S_sampleID_taxa_abundances.rds")
#microbiome_sampleID_taxa_abundances<-merge(microbiome_sampleID,genus, by="DADA_id")
#saveRDS(microbiome_sampleID_taxa_abundances,"../out/microbiome16S_sampleID_genus_abundances.rds")
#saveRDS(microbiome_sampleID_taxa_abundances,"../out/microbiome16S_sampleID_genus_abundances_scaled.rds")
############
## Phylum ##
############
phyla_list<-readRDS("../../1_data/out/Phyla_List.RDS")
phyla_list_over50percent<-phyla_list%>%
  filter(Samples_Containing_This_Phyla>=(18712*0.5))
phyla_over50percent<-phyla_list_over50percent$taxa_phyla
phyla_list_under50percent<-phyla_list%>%
  filter(Samples_Containing_This_Phyla<=(18712*0.5))%>%
  filter(Samples_Containing_This_Phyla>18712*0.02)
phyla_under50percent<-phyla_list_under50percent$taxa_phyla
phylum<-readRDS("../../1_data/out/phylum_prevalance_based.rds")
metadata<-merge(metadata,phylum, by="DADA_id")

############
## Family ##
############
family_list<-readRDS("../../1_data/out/family_List.RDS")
family_list_over50percent<-family_list%>%
  filter(Samples_Containing_This_Family>=(18712*0.5))
family_over50percent<-family_list_over50percent$taxa_family
family_list_under50percent<-family_list%>%
  filter(Samples_Containing_This_Family<(18712*0.5))%>%
  filter(Samples_Containing_This_Family>18712*0.02)%>%
  slice(-grep("Bacteria_Firmicutes_Bacillales_Thermoactinomycetaceae",taxa_family))
family_under50percent<-family_list_under50percent$taxa_family

family<-readRDS("../../1_data/out/family_prevalance_based.rds")
metadata<-merge(metadata,family, by="DADA_id")

###########
## Genus ##
###########
genus_raw<-readRDS("../../1_data/out/reads_genus.rds")
genus<-genus_raw
genus$DADA_id<-rownames(genus)
metadata<-merge(metadata,genus, by="DADA_id")

genus_list<-readRDS("../../1_data/out/Genus_List.RDS")
genus_75percent<-genus_list%>%
  filter(Samples_Containing_This_Genus>=(18712*0.75))
genus_over75percent<-genus_75percent$taxa_genus

genus_under75percent_list<-genus_list%>%
  filter(Samples_Containing_This_Genus<(18712*0.75))%>%
  filter(Samples_Containing_This_Genus>18712*0.02)
genus_under75percent<-genus_under75percent_list$taxa_genus

################
## Ordination ##
################
ordination_raw<-readRDS("../../2_beta/out/ordination_data_bstatus0.rds")
ordination<-ordination_raw%>%
  select(-c(contains("NMDS")))
colnames(ordination)[2:ncol(ordination)] = paste0("betadiv_",colnames(ordination)[2:ncol(ordination)])
betadiv_names<-colnames(ordination)[2:ncol(ordination)]
metadata<-merge(metadata,ordination, by="DADA_id")
metadata$plate<-as.factor(metadata$plate)

metadata_M<-metadata%>%
  filter(sex=='M')%>% #5259
  drop_na(diet_PC1) #4972
metadata_F<-metadata%>%
  filter(sex=='F')%>% #8289
  drop_na(diet_PC1) #7874

repstate_raw<-readRDS("../../1_data/out/reprostate_raw.rds")
repstate<-repstate_raw%>%
    dplyr::select(tid,state)

metadata_F<-merge(metadata_F,repstate, by="tid")%>%
  mutate(state=case_when(state=="C"~"Cycling",
                         state=="P"~"Pregnant",
                         state=="L"~"Lactating",
                         is.na(state)&sex=="F"&matstate=="juvenile"~"Immature"))
metadata_F$repro<-paste(metadata_F$matstate, metadata_F$state, sep = "_")
beep(2)
#saveRDS(metadata_all, "../../1_data/out/metadata_all_diet_grp_rain_hormone_reads_bstatus_less_than_6month_zscored.rds")
```

To understand how different microbial phenotypes change with age, we're taking a model comparison approach. For each phenotype, I've run the model below and pulled out the estimate, p-value, and AIC for age (and, if applicable, age^2^):

* Fixed Effects
    + Read count 
    + Age
    + First 5 Diet PCs
    + Rain 1 months prior to sample collection to account for fresh water vs watering holes
    + Rain 1 year prior to sample collection as an estimation of environment quality
    + Maturation state
    + *Quadratic Models Only*: 
        - Age^2^
    + *Female Models Only*
        - Reproductive State (matstate)
* Random effects
    + Group ID (grp_name)
    + Sample plate (plate)
    + Individual ID (sname)
    

#Males {.tabset .tabset-pills}
##Linear Models {.tabset}
```{r M_linear_basic}
phenotype<-c("richness","ShannonH","Simpson","Hill.q1","Hill.q2")
phenotype<-c(phenotype,betadiv_names,phyla_over50percent, family_over50percent,genus_over75percent)

M_linear_basic<-lapply(phenotype, function(x){
  print(x)
  lmer(substitute(i~age.years+
                    read_count.z+
                    rain_1moprior.z+
                    rain_1yrprior.z+
                    diet_PC1.z+
                    diet_PC2.z+
                    diet_PC3.z+
                    diet_PC4.z+
                    diet_PC5.z+
                    (1|sname)+(1|plate)+(1|grp_name),
              list(i=as.name(x))),data=metadata_M)
}) 
summaries_M_linear_basic<-lapply(M_linear_basic, summary)
AIC_M_linear_basic_list<-lapply(M_linear_basic,AIC)

Estimate_M_linear_basic<-vector(mode="numeric",length=length(phenotype))
SE_M_linear_basic<-vector(mode="numeric",length=length(phenotype))
P_value_M_linear_basic<-vector(mode="numeric",length=length(phenotype))
AIC_M_linear_basic<-vector(mode="numeric",length=length(phenotype))
distribution_type<-rep("gaussian",times=length(phenotype))

for (i in 1:length(phenotype)){
summaries2<-map(summaries_M_linear_basic,"coefficients")
summary<-as.data.frame(summaries2[[i]])
Estimate_M_linear_basic[i]<-summary[2,1]
SE_M_linear_basic[i]<-summary[2,2]
P_value_M_linear_basic[i]<-summary[2,5]
AIC_M_linear_basic[i]<-AIC_M_linear_basic_list[[i]]
}
age_linear_basic<-as.data.frame(cbind(phenotype,
                                      Estimate_M_linear_basic,
                                      SE_M_linear_basic,
                                      P_value_M_linear_basic,
                                      distribution_type))
colnames(age_linear_basic)[colnames(age_linear_basic)=="Estimate_M_linear_basic"] <- "Estimate";
colnames(age_linear_basic)[colnames(age_linear_basic)=="P_value_M_linear_basic"] <- "p_value";
colnames(age_linear_basic)[colnames(age_linear_basic)=="SE_M_linear_basic"] <- "SE";

saveRDS(age_linear_basic,"../out/ModelComparison_male_basic_linear.rds")
```


##Quadratic Models {.tabset }

```{r M_quad_basic, eval=F}
metadata_M$age.squared<-(as.numeric(metadata_M$age.years))^2

M_quad_basic<-lapply(phenotype, function(x){
  lmer(substitute(i~age.years+
                    age.squared+
                    read_count.z+
                    rain_1moprior.z+
                    rain_1yrprior.z+
                    diet_PC1.z+
                    diet_PC2.z+
                    diet_PC3.z+
                    diet_PC4.z+
                    diet_PC5.z+
                    (1|sname)+(1|plate)+(1|grp_name),
              list(i=as.name(x))),data=metadata_M)
})
summaries_M_quad_basic<-lapply(M_quad_basic, summary)
AIC_M_quad_basic_list<-lapply(M_quad_basic,AIC)

Estimate_M_quad_basic<-vector(mode="numeric",length=length(phenotype))
P_value_M_quad_basic<-vector(mode="numeric",length=length(phenotype))
Term_A<-vector(mode="character",length=length(phenotype))
Term_B<-vector(mode="character",length=length(phenotype))
SE_M_quad_basic<-vector(mode="character",length=length(phenotype))
AIC_M_quad_basic<-vector(mode="numeric",length=length(phenotype))
distribution_type<-rep("gaussian",times=length(phenotype))

for (i in 1:length(phenotype)){
summaries2<-map(summaries_M_quad_basic,"coefficients")
summary<-as.data.frame(summaries2[[i]])
Estimate_M_quad_basic[i]<-summary[2,1]
P_value_M_quad_basic[i]<-summary[2,5]
rownames<-rownames(summary)
Term_A[i]<-rownames[2]
SE_M_quad_basic[i]<-summary[2,2]
AIC_M_quad_basic[i]<-AIC_M_quad_basic_list[[i]]
}
M_quad_basic_age<-as.data.frame(cbind(Term_A,
                                      phenotype,
                                      Estimate_M_quad_basic,
                                      SE_M_quad_basic,
                                      P_value_M_quad_basic,
                                      distribution_type))
colnames(M_quad_basic_age)[colnames(M_quad_basic_age)=="Estimate_M_quad_basic"] <- "Estimate_A";
colnames(M_quad_basic_age)[colnames(M_quad_basic_age)=="P_value_M_quad_basic"] <- "p_value_A";
colnames(M_quad_basic_age)[colnames(M_quad_basic_age)=="SE_M_quad_basic"] <- "SE_A";

for (j in 1:length(phenotype)){
summaries2<-map(summaries_M_quad_basic,"coefficients")
summary<-as.data.frame(summaries2[[j]])
Estimate_M_quad_basic[j]<-summary[3,1]
P_value_M_quad_basic[j]<-summary[3,5]
SE_M_quad_basic[i]<-summary[3,2]
rownames<-rownames(summary)
Term_B[j]<-rownames[3]
}
M_quad_basic_age.sq<-as.data.frame(cbind(Term_B,
                                         phenotype,
                                         Estimate_M_quad_basic,
                                         SE_M_quad_basic,
                                         P_value_M_quad_basic,
                                         distribution_type))
colnames(M_quad_basic_age.sq)[colnames(M_quad_basic_age.sq)=="Estimate_M_quad_basic"] <- "Estimate_B";
colnames(M_quad_basic_age.sq)[colnames(M_quad_basic_age.sq)=="P_value_M_quad_basic"] <- "p_value_B";
colnames(M_quad_basic_age.sq)[colnames(M_quad_basic_age.sq)=="SE_M_quad_basic"] <- "SE_B";

M_quad_basic<-merge(M_quad_basic_age,M_quad_basic_age.sq, by=c("phenotype","distribution_type"))
saveRDS(M_quad_basic,"Ch1/out/ModelComparison_male_basic_quad1.rds")
```

#Females {.tabset .tabset-pills}
##Linear Models {.tabset }

```{r F_linear_basic}

F_linear_basic<-lapply(phenotype, function(x){
  lmer(substitute(i~age.years+
                    read_count+
                    rain_1moprior+
                    rain_1yrprior+
                    matstate+
                    diet_PC1+
                    diet_PC2+
                    diet_PC3+
                    diet_PC4+
                    diet_PC5+
                    (1|sname)+(1|plate)+(1|grp_name),
              list(i=as.name(x))),data=metadata_F)
}) 
summaries_F_linear_basic<-lapply(F_linear_basic, summary)
AIC_F_linear_basic_list<-lapply(F_linear_basic,AIC)

Estimate_F_linear_basic<-vector(mode="numeric",length=length(phenotype))
SE_F_linear_basic<-vector(mode="numeric",length=length(phenotype))
P_value_F_linear_basic<-vector(mode="numeric",length=length(phenotype))
AIC_F_linear_basic<-vector(mode="numeric",length=length(phenotype))

for (i in 1:length(phenotype)){
summaries2<-map(summaries_F_linear_basic,"coefficients")
summary<-as.data.frame(summaries2[[i]])
Estimate_F_linear_basic[i]<-summary[2,1]
SE_F_linear_basic[i]<-summary[2,2]
P_value_F_linear_basic[i]<-summary[2,5]
AIC_F_linear_basic[i]<-AIC_F_linear_basic_list[[i]]
}
age_linear_basic<-as.data.frame(cbind(phenotype,Estimate_F_linear_basic,SE_F_linear_basic,P_value_F_linear_basic))
colnames(age_linear_basic)[colnames(age_linear_basic)=="Estimate_F_linear_basic"] <- "Estimate";
colnames(age_linear_basic)[colnames(age_linear_basic)=="P_value_F_linear_basic"] <- "p_value";
```

## Quadratic Model

```{r F_quad_basic}
metadata_F$age.squared<-(as.numeric(metadata_F$age.years))^2

F_quad_basic<-lapply(phenotype, function(x){
  lmer(substitute(i~age.years+
                    age.squared+
                    read_count+
                    rain_1moprior+
                    rain_1yrprior+
                    repro+
                    diet_PC1+
                    diet_PC2+
                    diet_PC3+
                    diet_PC4+
                    diet_PC5+
                    (1|sname)+(1|plate)+(1|grp_name),
              list(i=as.name(x))),data=metadata_F)
})
summaries_F_quad_basic<-lapply(F_quad_basic, summary)
AIC_F_quad_basic_list<-lapply(F_quad_basic,AIC)

Estimate_F_quad_basic<-vector(mode="numeric",length=length(phenotype))
P_value_F_quad_basic<-vector(mode="numeric",length=length(phenotype))
Term<-vector(mode="character",length=length(phenotype))
AIC_F_quad_basic<-vector(mode="numeric",length=length(phenotype))

for (i in 1:length(phenotype)){
summaries2<-map(summaries_F_quad_basic,"coefficients")
summary<-as.data.frame(summaries2[[i]])
Estimate_F_quad_basic[i]<-summary[2,1]
P_value_F_quad_basic[i]<-summary[2,5]
rownames<-rownames(summary)
Term[i]<-rownames[2]
AIC_F_quad_basic[i]<-AIC_F_quad_basic_list[[i]]
}
age_quad_basic_age<-as.data.frame(cbind(Term_A,phenotype,Estimate_F_quad_basic,P_value_F_quad_basic))
colnames(age_quad_basic_age)[colnames(age_quad_basic_age)=="Estimate_F_quad_basic"] <- "Estimate_A";
colnames(age_quad_basic_age)[colnames(age_quad_basic_age)=="P_value_F_quad_basic"] <- "p_value_A";


for (j in 1:length(phenotype)){
summaries2<-map(summaries_F_quad_basic,"coefficients")
summary<-as.data.frame(summaries2[[j]])
Estimate_F_quad_basic[j]<-summary[3,1]
P_value_F_quad_basic[j]<-summary[3,5]
rownames<-rownames(summary)
Term[j]<-rownames[3]
}
age_quad_basic_age.sq<-as.data.frame(cbind(Term_B,phenotype,Estimate_F_quad_basic,P_value_F_quad_basic))
colnames(age_quad_basic_age.sq)[colnames(age_quad_basic_age.sq)=="Estimate_F_quad_basic"] <- "Estimate_B";
colnames(age_quad_basic_age.sq)[colnames(age_quad_basic_age.sq)=="P_value_F_quad_basic"] <- "p_value_B";

age_quad_basic_F<-merge(age_quad_basic_age,age_quad_basic_age.sq, by="phenotype")
```

#Relative Abundance Plots
```{r fig.height=8,fig.width=16}
relab_phylum_long<-readRDS("../../1_data/out/relabundance_long_phyla_nochl.rds") %>% 
  filter(!grepl("Chloroplast",taxa_phyla)) %>% 
  filter(taxa_phyla!="Bacteria_Cyanobacteria") %>% 
  group_by(sample_ID) %>% 
  mutate(low_quantity=if_else(rel_abundance<0.01,T,F),#grouping the ones under 1%
         total_reads=sum(count)) %>% 
  ungroup() %>% 
  mutate(taxa_phyla=gsub(pattern = "_",replacement = "/",taxa_phyla))

#bring in the metadata
metadata_random<-readRDS("../../4_machinelearning/out/metadata_random.rds") %>% 
  mutate(age_matured=as.numeric((matured-birth)/365.25)) %>% 
  filter(collect_grp<3) %>% 
  select(DADA_id,sname,age.years, statdate, status,birth,sex) %>% 
  rename(age_chrono=age.years,
         sample_ID=DADA_id) 

#smoosh the tables together and group into each year by sex
relab_phylum_long2<-relab_phylum_long %>% 
  left_join(metadata_random,by="sample_ID") %>% 
  drop_na() %>% 
  mutate(age_floor=floor(age_chrono)) %>% 
  group_by(taxa_phyla,sex,age_floor) %>% 
  summarise(mean_relabun=mean(rel_abundance)) %>% 
  mutate(plot_taxa = ifelse(mean_relabun<0.01,"Low Abundance Taxa (<1%)", taxa_phyla)) %>%
  ungroup() %>% 
  mutate(plot_taxa = ifelse(plot_taxa == "Bacteria/NA" | plot_taxa == "NA/NA", 
                            "Unclassified Taxa", 
                            plot_taxa)) %>% 
  group_by(plot_taxa, sex, age_floor) %>% 
  summarise(mean_relabun = sum(mean_relabun, na.rm = T), .groups = "drop") %>% 
  mutate(plot_taxa=gsub("[A-z]+/","",plot_taxa)) 

#females
p2a_relabunF<-relab_phylum_long2 %>% 
  filter(sex=="F") %>% 
  ggplot(aes(x=age_floor,y=mean_relabun,fill=plot_taxa),width=1)+
  geom_col(color="grey40")+
  labs(x="Age (years)",y="Mean Relative \n Abundance",fill='Phylum',title="Females",tag="A")+
  theme_classic()+
  scale_x_continuous(expand=c(0,0))+
  scale_y_continuous(expand=c(0,0))+
  scale_fill_viridis_d(option = "turbo",name=NULL)+
 theme(axis.text = element_text(size = 32, colour = "black"),
        axis.title=element_text(size=39),
        legend.text = element_text(size=30),
        plot.tag=element_text(size=44),
        plot.title=element_text(size=42),
        legend.position = "none");p2a_relabunF
ggsave(plot=p2a_relabunF, filename = "../out/Figures/RelAbunPhylumF.png",width = 16, height=8, units="in",dpi=300)

#males
p2b_relabunM<-relab_phylum_long2 %>% 
  filter(sex=="M") %>% 
  ggplot(aes(x=age_floor,y=mean_relabun,fill=plot_taxa),width=1)+
  geom_col(color="grey30")+
  labs(x=NULL,y=NULL,fill='Phylum')+
  labs(x="Age (years)",y="Mean Relative \n Abundance",fill='Phylum',title="Males",tag="B")+
  theme_classic()+
  scale_x_continuous(expand=c(0,0))+
  scale_y_continuous(expand=c(0,0))+
  scale_fill_viridis_d(option = "turbo",name=NULL)+
  theme(axis.text = element_text(size = 32, colour = "black"),
        axis.title=element_text(size=39),
        legend.text = element_text(size=34),
        plot.tag=element_text(size=44),
        plot.title=element_text(size=42),
        legend.position = "right",
        legend.key.width=unit(1,"cm"));p2b_relabunM
ggsave(plot=p2b_relabunM, filename = "../out/Figures/RelAbunPhylumM.png",width = 18, height=8, units="in",dpi=300)
```

``````{r fig.height=10,fig.width=35}
library(ggpubr)
ggarrange(p2A, p2B, ncol=2, nrow=1)   
ggsave("../out/Figures/2AB_RelAbunPhylum.png")
###
```

#Barplots
```{r fig.height=6,fig.width=15}
age_gaussian_quad<-readRDS("../../5_feature-analyses/out/FeatureAnalysis_LMM_L1O_gaussian_quad_meancentered_May21.rds")
#age_gaussian_quad<-readRDS("../../5_feature-analyses/out/FeatureAnalysis_LMM_L1O_gaussian_quad_inclASVs25to50.rds")
colnames(age_gaussian_quad)<-gsub("_A","_linear",colnames(age_gaussian_quad));
colnames(age_gaussian_quad)<-gsub("_B","_quad",colnames(age_gaussian_quad));
colnames(age_gaussian_quad)<-gsub("Term","Age",colnames(age_gaussian_quad));

gaussian_features<-as.data.frame(readRDS("../../4_machinelearning/out/phenotype_types.rds")) %>% 
  mutate(Feature=as.character(phenotype_names)) %>% select(-phenotype_names)
gaussian_features2<-gaussian_features
gaussian_features2[gaussian_features2=="ASV"]<-"Gaussian ASV"

gaussian<-age_gaussian_quad %>% 
  left_join(gaussian_features2,by="Feature") %>% 
  rename(Feature_Category=phenotype_type) %>% 
  mutate_if(is.factor, as.character)%>%
  mutate_at(vars(matches("(Estimate)|(SE)|(p_value)")), as.numeric) %>% 
  mutate(p_linear_adj=p.adjust(p_value_linear, method="BH"),
         p_quad_adj=p.adjust(p_value_quad, method="BH")) %>% 
  select(-matches("Age"))

binomial<-readRDS("../../5_feature-analyses/out/binomial_features_BHadjusted2_wide.rds") %>% 
  mutate(p_linear_adj=p.adjust(p_value_linear, method="BH"),
         p_quad_adj=p.adjust(p_value_quad, method="BH"))
all_features<-rbind(gaussian,binomial)

AllSigFeatures<-all_features %>% 
  mutate(Significance_linear = if_else(p_linear_adj<=0.05,T,F),
         Significance_quad = if_else(p_quad_adj<=0.05,T,F)) 

AllSigFeatures$Feature_Category[AllSigFeatures$Feature_Category=="beta_diversity"]<-"Composition"
AllSigFeatures$Feature_Category[AllSigFeatures$Feature_Category=="alpha_diversity"]<-"Alpha Diversity"
AllSigFeatures$Feature_Category[AllSigFeatures$Feature_Category=="phylum"]<-"Phylum"
AllSigFeatures$Feature_Category[AllSigFeatures$Feature_Category=="family"]<-"Family"
AllSigFeatures$Feature_Category[AllSigFeatures$Feature_Category=="genus"]<-"Genus"
feature_levels<-c("Alpha Diversity","Composition", "Phylum", "Family", "Genus", "Gaussian ASV", "Binomial ASV")
AllSigFeatures$Feature_Category<-factor(AllSigFeatures$Feature_Category, 
                                              levels=feature_levels)
GausSigFeatures<-AllSigFeatures %>% 
  filter(Feature_Category!="Binomial ASV") 

communitymetrics<-gaussian %>% 
  filter(Feature_Category=="alpha_diversity"|Feature_Category=="beta_diversity") %>% 
  select(Feature,Feature_Category,Estimate_linear,SE_linear,p_linear_adj,Estimate_quad,SE_quad,p_quad_adj) %>% 
  filter(p_linear_adj<=0.05|p_quad_adj<=0.05) #removed for rds output below
communitymetrics$Feature<-gsub("betadiv_PC","PC",communitymetrics$Feature)

#don't forget to rbind these to the rest for the plot
GausSigLinear<-GausSigFeatures %>% 
  arrange(Estimate_linear) %>% 
  slice(1:25,nrow(GausSigFeatures):(nrow(GausSigFeatures)-24)) %>% #removed for rds output below
  select(Feature,Feature_Category,Estimate_linear,SE_linear,p_linear_adj,Estimate_quad,SE_quad,p_quad_adj)

silva<-readRDS("../../1_data/out/ASVs_silva_assigned_extendedMay21.rds") %>% 
  select(ASV,ASV_id,starts_with("taxa_")) %>% 
  rename(Feature=ASV)

metricsgauss_linear<-rbind(communitymetrics,GausSigLinear) %>% 
  left_join(silva,by="Feature") %>% 
  mutate(taxa_lab=case_when(Feature=="Hill.q1"~"Hill Number 1",
                            Feature=="Hill.q2"~"Hill Number 2",
                            Feature=="richness"~"ASV Richness",
                            Feature=="ShannonH"~"Shannon's H",
                            Feature=="Simpson"~"Simpson's Diversity",
                            Feature_Category=="Gaussian ASV"~ASV_id,
                            Feature_Category!="Gaussian ASV"&
                              Feature_Category!="alpha_diversity"~Feature)) %>% 
  mutate(taxa_lab=if_else(p_quad_adj<=0.05,paste(taxa_lab,"*",sep=""),taxa_lab)) %>% 
  mutate(taxa_col=case_when(Feature_Category=="alpha_diversity"~ "#F8766D",
                            Feature_Category=="beta_diversity"~ "#C5B33E",
                            Feature_Category=="Phylum"~ "#00BA38",
                            Feature_Category=="Family"~ "#01BFC4",
                            Feature_Category=="Genus"~ "#629DFF",
                            Feature_Category=="Gaussian ASV"~ "#F564E3")) 

metricsgauss_linear$Feature_Category<-factor(metricsgauss_linear$Feature_Category,
                                      levels=c("alpha_diversity","beta_diversity",
                                               "Phylum","Family","Genus","Gaussian ASV"))
#saveRDS(metricsgauss_linear,"../out/GaussianLMMOutputClean.rds")
metricsgauss_linear<-metricsgauss_linear %>% 
  arrange(Feature_Category,Estimate_linear) %>% 
  mutate(taxa_lab2=case_when(taxa_lab=="Bacteria_Cyanobacteria_Gastranaerophilales_NA*"~
                               "UC Family in Gastranaerophilales*",
                             taxa_lab=="Bacteria_Firmicutes_NA_NA"~
                               "UC Family in Firmicutes",
                             taxa_lab=="Bacteria_Cyanobacteria_Gastranaerophilales_NA_NA*"~
                               "UC Genus in Gastranaerophilales*",
                             taxa_lab=="Bacteria_Firmicutes_Clostridiales_Peptostreptococcaceae_NA"~
                               "UC Genus in Peptostreptococcaceae",
                             taxa_lab=="Bacteria_Firmicutes_NA_NA_NA"~
                               "UC Genus in Firmicutes",
                             taxa_lab=="Bacteria_Proteobacteria_Enterobacteriales_Enterobacteriaceae_NA (G)*"~
                               "UC Genus in Enterobacteriaceae*",
                             taxa_lab=="Bacteria_Firmicutes_Clostridiales_Clostridiaceae 1_NA*"~
                               "UC Genus in Clostridiaceae 1*",
                             taxa_lab=="Bacteria_Firmicutes_Clostridiales_Clostridiaceae 1_Clostridium sensu stricto 1"~
                               "Clostridium sensu stricto 1")) %>% 
  mutate(taxa_lab2=if_else(is.na(taxa_lab2),taxa_lab,taxa_lab2)) %>% 
  mutate(taxa_lab2=gsub("[A-z]+_","",taxa_lab2)) %>% 
  select(Feature,Feature_Category,Estimate_linear,SE_linear,p_linear_adj,Estimate_quad,SE_quad,p_quad_adj,taxa_lab,taxa_col,taxa_lab2) 

metricsgauss_linear$no<-c(1:nrow(metricsgauss_linear))
#metricsgauss$no<-c(1:6,15,7:14,16:nrow(metricsgauss)) when organizing by Feature alphabetically

#metricsgauss_linear$taxa_lab <- gsub("_", ">", metricsgauss_linear$taxa_lab)

communitymetrics<-metricsgauss_linear %>% filter(Feature_Category=="alpha_diversity"| 
                                      Feature_Category=="beta_diversity")
taxametrics<-metricsgauss_linear %>% filter(!Feature_Category=="alpha_diversity"& 
                                      !Feature_Category=="beta_diversity")


p2C_communityestimates <- ggplot(communitymetrics, aes(x=rev(no), y=Estimate_linear)) +
  geom_bar(stat="identity",fill=communitymetrics$taxa_col,width=.75)+
  geom_hline(yintercept=0,linetype="dashed",color="grey")+
  scale_x_discrete(limits = communitymetrics$no, 
                   labels = rev(communitymetrics$taxa_lab))+
  geom_errorbar(aes(ymax = Estimate_linear+SE_linear, 
                   ymin = Estimate_linear-SE_linear), 
                   colour=communitymetrics$taxa_col, size=0.6,width=0.5)+
  labs(y = "Linear Estimate",x="")+theme_bw()+
  theme(legend.position = "none",
        axis.text = element_text(size=30,color="black"),
        axis.title = element_text(size=34),
        plot.tag=element_text(size=44),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())+
  coord_flip()+labs(tag="C");p2C_communityestimates
ggsave(plot=p2C_communityestimates,filename = "../out/Figures/2C_CommunityEstimates.png",height=6,width=15,dpi=300,units="in")
```

```{r fig.height=21,fig.width=18}

p2E_gausestimates <- ggplot(taxametrics, aes(x=rev(no), y=Estimate_linear)) +
  geom_bar(stat="identity",fill=taxametrics$taxa_col,width=.75)+
  geom_hline(yintercept=0,linetype="dashed",color="grey")+
  scale_x_discrete(limits = taxametrics$no, 
                   labels = c(rev(taxametrics$taxa_lab2),function(x) str_wrap(x, width = 10)))+
  #scale_x_discrete(labels = function(x) str_wrap(x, width = 20)) +
  geom_errorbar(aes(ymax = Estimate_linear+SE_linear, 
                    ymin = Estimate_linear-SE_linear), 
                    colour=taxametrics$taxa_col, size=0.6,width=0.5) +
  labs(y = "Linear Estimate",x="")+theme_bw()+coord_flip()+labs(tag="D")+
  theme(legend.position = "bottom",
        axis.text = element_text(size=32,color="black"),
        axis.title = element_text(size=37,color="black"),
        plot.tag=element_text(size=44),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()); p2E_gausestimates
ggsave(plot=p2E_gausestimates,filename = "../out/Figures/2E_Gaussian50Estimates.png", height=21,width=18,units="in", dpi=300)
```

# Facet Plots
```{r fig.height=12,fig.width=22}
ordination<-readRDS("../../2_beta/out/ordination_data_bstatus0.rds") %>% 
  select(DADA_id, starts_with("PC")) %>% 
  select(-c(PC3,PC9,PC10))
colnames(ordination)<-c("DADA_id",paste("PC ",seq(1:2),"*",sep=""),paste("PC ",4:8,"*",sep=""))

cow_metadata<-readRDS("../../4_machinelearning/out/metadata_bstatus0_mathormonerank_rainpriortomilestone_matrankbirthgroup.rds") %>% 
  rename(age_chrono=age.years) %>%
  select(DADA_id,age_chrono,ShannonH,Simpson,Hill.q1,Hill.q2) %>% 
  rename("Hill No. 1*"=Hill.q1,
         "Hill No. 2*"=Hill.q2,
         "Shannon H*"=ShannonH,
         "Simpson's Diversity*"=Simpson) %>% 
  left_join(ordination,by="DADA_id") %>% 
  mutate(age_floor=floor(age_chrono))

most<-cow_metadata%>%
  select(-DADA_id,-age_chrono)%>%
  gather(key = 'Feature', value = 'value', -age_floor) %>%
  group_by(age_floor, Feature) %>%
  summarise(age_value = mean(value)) %>% 
  ungroup() %>% 
  mutate(color = ifelse(grepl("PC", Feature), "#C5B33E", "#F8766D")) 
most2<-most %>%
  group_by(Feature) %>% 
  summarise() %>% 
  ungroup() 
most2$no<-c(1:2,5:11,3:4)
#most2$no<-c(5:6,11,9:10,8,7,1:4)#setting the sort order
most2<-arrange(most2,no)
names<-most2$Feature
most$Feature<-factor(most$Feature,levels=names)

p2D_communityage<-ggplot(data=most,
            aes(x=age_floor,
                y=age_value))+
  geom_line(stat="smooth",method = "loess", size = 2,alpha = 0.5,aes(col=color))+
  geom_point(aes(group=age_floor,col=color),size=4)+theme_bw()+
  scale_color_manual(values=c("#C5B33E","#F8766D"))+labs(tag="E")+
  facet_wrap(~Feature, scales = "free",ncol=4,
             labeller = label_wrap_gen(width=20,multi_line = TRUE))+
  theme(strip.text.x = element_text(size = 32),
        axis.text = element_text(color="black", size=27),
        axis.title = element_text(color="black", size=36),
        plot.tag=element_text(size=42),
        legend.position = "none",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())+
  labs(x="Age (years)",
       y="Average Value");p2D_communityage
ggsave(plot=p2D_communityage, filename = "../out/Figures/2D_CommunityMetricsAge.png",height=12,width=22,units="in",dpi=300)
```

```{r fig.height=20,fig.width=22}
silva<-readRDS("../../1_data/out/ASVs_silva_assigned_extendedMay21.rds") %>% 
  select(ASV,ASV_id,starts_with("taxa_")) %>% 
  rename(Feature=ASV)
silva2<-silva %>% select(Feature,ASV_id)

top50features<-metricsgauss_linear %>% 
  filter(Feature_Category!="alpha_diversity"&Feature_Category!="beta_diversity") %>% 
  arrange(Estimate_linear) 

top50features_list<-top50features$Feature
phenotypes_readcount_raw<-readRDS('../../4_machinelearning/out/phenotypes_readcounts_no_chl_mt.rds')
phenotypes_readcount<-phenotypes_readcount_raw%>%
  column_to_rownames("DADA_id")
phenotypes_readcount[phenotypes_readcount>0]<-1

duck_metadata<-readRDS("../../4_machinelearning/out/metadata_bstatus0_mathormonerank_rainpriortomilestone_matrankbirthgroup.rds") %>% 
  rename(age_chrono=age.years) %>% select(DADA_id,age_chrono)

filtered_phenotypes_top50<-phenotypes_readcount%>%
  select(.dots=top50features_list) %>% 
  rownames_to_column("DADA_id")
top50all_phenotype_names<-c("DADA_id",top50features_list)
colnames(filtered_phenotypes_top50)<-top50all_phenotype_names

metadata_top50all<-merge(duck_metadata,filtered_phenotypes_top50, by="DADA_id")
metadata_top50all$age.floor<-floor(metadata_top50all$age_chrono)

labels<-metricsgauss_linear %>% 
  select(Feature,taxa_lab,no,Feature_Category) %>% 
  mutate(taxa_lab2=case_when(taxa_lab=="Bacteria_Cyanobacteria_Gastranaerophilales_NA*"~
                               "UC Family in Gastranaerophilales*",
                             taxa_lab=="Bacteria_Firmicutes_NA_NA"~
                               "UC Family in Firmicutes",
                             taxa_lab=="Bacteria_Cyanobacteria_Gastranaerophilales_NA_NA*"~
                               "UC Genus in Gastranaerophilales*",
                             taxa_lab=="Bacteria_Firmicutes_Clostridiales_Peptostreptococcaceae_NA"~
                               "UC Genus in Peptostreptococcaceae",
                             taxa_lab=="Bacteria_Firmicutes_NA_NA_NA"~
                               "UC Genus in Firmicutes",
                             taxa_lab=="Bacteria_Proteobacteria_Enterobacteriales_Enterobacteriaceae_NA*"~
                               "UC Genus in Enterobacteriaceae*",
                             taxa_lab=="Bacteria_Firmicutes_Clostridiales_Clostridiaceae 1_NA*"~
                               "UC Genus in Clostridiaceae 1*",
                             taxa_lab=="Bacteria_Firmicutes_Clostridiales_Clostridiaceae 1_Clostridium sensu stricto 1"~
                               "Clostridium sensu stricto 1")) %>% 
  mutate(taxa_lab2=if_else(is.na(taxa_lab2),taxa_lab,taxa_lab2)) %>% 
  mutate(taxa_lab2=gsub("[A-z]+_","",taxa_lab2)) 

most<-metadata_top50all%>%
  select(age.floor,top50all_phenotype_names)%>%
  select(-DADA_id)%>%
  gather(key = 'Feature', value = 'pres_abs', -age.floor) %>%
  group_by(age.floor, Feature) %>%
  summarise(total_samples = n(),
            taxa_in_sample = sum(pres_abs),
            prevalence=taxa_in_sample/total_samples) %>% 
  ungroup() %>% 
  left_join(labels,by="Feature") %>% 
  left_join(silva2,by="Feature") %>% 
  mutate(Feature=if_else(Feature_Category=="Gaussian ASV",ASV_id,Feature)) %>% 
  arrange(Feature_Category) %>% 
    mutate(taxa_col=case_when(Feature_Category=="Phylum"~ "#00BA38",
                            Feature_Category=="Family"~ "#01BFC4",
                            Feature_Category=="Genus"~ "#629DFF",
                            Feature_Category=="Gaussian ASV"~ "#F564E3")) %>% 
  filter(Feature_Category!="Gaussian ASV")

most2<-most %>%
  group_by(Feature,taxa_lab2,no) %>% 
  summarise() %>% 
  ungroup() %>% 
  arrange(no) %>% 
  mutate(taxa_lab2=as.factor(taxa_lab2))
taxa_labels<-most2$taxa_lab2


most$Feature=factor(most$taxa_lab2, levels=most2$taxa_lab2)

p2F_gausage<-ggplot(data=most,
            aes(x=age.floor,
                y=prevalence,color=taxa_col))+
  geom_line(stat="smooth",method = "loess", size = 2,alpha = 0.5,aes(col=taxa_col))+
  geom_point(aes(group=age.floor),size=4)+theme_bw()+
  scale_color_manual(values=c("#00BA38","#01BFC4","#629DFF","#F564E3"))+
  labs(tag="F")+
  facet_wrap(~Feature, scales = "free",ncol=4,
             labeller = label_wrap_gen(width=20,multi_line = TRUE))+
  theme(strip.text.x = element_text(size = 32),
        axis.text = element_text(color="black", size=27),
        axis.title = element_text(color="black", size=37),
        plot.tag=element_text(size=42),
        legend.position = "none",
        panel.grid = element_blank())+
  labs(x="Age (years)",
       y="Prevalance");p2F_gausage
ggsave(plot=p2F_gausage,filename = "../out/Figures/S3_Top50GausFeatures_noASV.png",height=20,width=22,dpi=300, units="in")
```

# Combined Plot
```{r}
cow=ggdraw()+
  draw_plot(p2a_relabunF, x=0, y=.8, width=.48,height = 0.18)+
  draw_plot(p2b_relabunM, x=.48, y=.8, width=.52,height=0.18)+
  draw_plot(p2C_communityestimates, x=0, y=.6, width=.4,height=0.2)+
  draw_plot(p2D_communityage, x=.4, y=.5, width=.6,height=0.3)+
  draw_plot(p2E_gausestimates, x=0, y=0, width=.4,height=0.6)+ #bottom left
  draw_plot(p2F_gausage, x=.4, y=0, width=.6,height=0.5) #bottom right

cow
  
ggsave(plot=cow,filename="../out/Figures/Fig2CowTest.png",width = 40, height = 35,units="in")

bull=ggdraw()+
  draw_plot(p2a_relabunF, x=0, y=.84, width=.5,height = 0.16)+
  draw_plot(p2b_relabunM, x=0, y=.68, width=.5,height=0.16)+
  draw_plot(p2C_communityestimates, x=0, y=.54, width=.5,height=0.14)+
  draw_plot(p2E_gausestimates, x=0, y=0, width=.5,height=0.54)+ #bottom left
  draw_plot(p2D_communityage, x=.50, y=.7, width=.5,height=0.3)+
  draw_plot(p2F_gausage, x=.50, y=0, width=.5,height=0.7) #bottom right

bull
  
ggsave(plot=bull,filename="../out/Figures/Fig2BullTest2.png",width = 38, height = 40,units="in")
```
