---
title: "SocioEnviroPredictors_StatBreakpoint"
author: "Mauna Dasari"
date: "2/14/2021"
output:
  html_document:
    number_sections: no
    theme: paper
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F,warning=F,message=F,cache=2)
library(feather);library(tidyverse);library(kableExtra);library(gridExtra);

library(lme4);library(lmerTest);library(mgcv);library(survival);library(survminer);
statuses<-readRDS("../../1_data/out/statuses_updatedDec2020.rds")
mom_ranks<-readRDS("../../1_data/out/mom_rank_updatedSep2021.rds") %>% select(sname,mom_rank)
metadata_all<-readRDS("../../4_machinelearning/out/metadata_bstatus0_mathormonerank_rainpriortomilestone_matrankbirthgroup.rds") %>% 
  rename(age_firstlivebirth=age_first_live_birth,
         age_firstpreg=age_first_preg,
         age_chrono=age.years) %>% 
  mutate(age_stat=(statdate-birth)/365.25) %>% 
  filter(collect_grp<3.0) %>% 
  select(-plate,-statdate,-status,-mom_rank,-read_count,-richness,-ShannonH,-Simpson,-starts_with("Hill"),-ordrank,-proprank) %>% 
  left_join(statuses,by="sname") %>% 
  left_join(mom_ranks,by="sname") 
```

This file will use median age of sexual maturity to determine the breakpoint between phases.

# What makes some animals have older looking microbiome or younger looking microbiomes?

## What predicts delta age? {.tabset}
$\Delta$age = $age_{microbial}$ - $age_{chronological}$, so if $\Delta$age is negative, that means $age_{chronological}$ > $age_{microbial}$ and the animal is microbially young for age. 

### Delta Over Life {.tabset}
````{r fig.height=10}
grp_adf<-readRDS("../../1_data/out/collectgrpdensity_adf.rds") %>% select(sname,collect_grp,collection_date,adf_collect)
metadata_simple<-metadata_all %>% select(DADA_id,sname,collection_date,season,collect_grp,rain_1yrprior) %>% 
  left_join(grp_adf, by=c("sname", "collection_date", "collect_grp")) %>% 
  mutate(collect_grp=as.factor(collect_grp)) %>% distinct()

monthly_temp_rain<-readRDS("../../4_machinelearning/out/monthly_temp_rain_hydroyear.rds") %>% 
  distinct() %>% 
  mutate(hydroyear=as.factor(hydroyear)) 

H.PredLife_raw<-readRDS("../../4_machinelearning/out/GaussianProcessModeling/H.PredictedAgeData.rds") 

PredLifeCov<-H.PredLife_raw %>% 
  select(-c(testID,delta_abs,starts_with("resid"))) %>% 
  left_join(metadata_simple, by=c("DADA_id","sname")) %>% 
  left_join(monthly_temp_rain,by=c("sname","collection_date")) 
PredLifeCov$collect_month<-as.factor(PredLifeCov$collect_month)
#saveRDS(PredLifeCov,"../out/GaussianProcessModeling/out/PredictedAge_EnviroCov.rds")

DSI<-readRDS("../../4_machinelearning/out/metadata_with_DSI.rds") %>% 
  select(DADA_id,sname,collect_grp,collection_date, DSI_F,DSI_M)
DSI$collect_grp<-as.factor(DSI$collect_grp)

EA_raw<-read.csv("../../4_machinelearning/in/ea_dataset_for_mauna.csv",header = T) %>% select(-X) 
EA<-EA_raw %>% 
  filter(included_cases=="TRUE") %>% 
  mutate(highGrpSize=case_when(group_size>=quantile(EA_raw$group_size, probs=0.75, na.rm=TRUE)~T,
                           group_size<quantile(EA_raw$group_size, probs=0.75, na.rm=TRUE)~F),
         lowMatSCI=case_when(maternal_SCI_F<=quantile(EA_raw$maternal_SCI_F, probs=0.25, na.rm=TRUE)~T,
                             maternal_SCI_F>quantile(EA_raw$maternal_SCI_F, probs=0.25, na.rm=TRUE)~F),
         lowMatRank=case_when(maternal_rank>=quantile(EA_raw$maternal_rank, probs=0.75, na.rm=TRUE)~T,
                           maternal_rank<quantile(EA_raw$maternal_rank, probs=0.75, na.rm=TRUE)~F)) %>% 
  mutate(cumulative_adversity=(highGrpSize+lowMatSCI+lowMatRank+sibling+maternal_loss+drought)) %>% 
  select(-c(mom,bstatus,status,statdate,group_size,
            maternal_rank,maternal_SCI_F,ea_check,included_cases)) %>% 
  mutate(EA=case_when(cumulative_adversity>=3~T,
                      cumulative_adversity<3~F)) #%>% 
  #select(sname,cumulative_adversity)
#saveRDS(EA,"../out/ea_dataset_cumulativeadversity.rds") #only sname and cumulative adversity
#saveRDS(EA,"../out/ea_dataset.rds")

ranks<-readRDS("../../1_data/out/rank_raw_adult_Feb2021.rds") %>% 
  select(sname, collection_date, ordrank,proprank)

PredLife.B<-PredLifeCov %>% 
  left_join(ranks,by=c("sname","collection_date")) %>% 
  left_join(DSI,by=c("sname","DADA_id","collection_date","collect_grp")) %>% 
  left_join(EA,by=c("sname","sex")) 

#write.csv(PredLife.B,"../out/milestone_data_verification/socioenviro_ageaccel_lifespan.csv")
```

#### Males {.tabset}
##### Sources of Adversity {.tabset}
```{r}
PredLife_M.B<-PredLife.B %>% filter(sex=="M")
model1<-lmer(delta~age_chrono+
               avg_month_maxtemp+
               sum_month_rain+
               season+
               maternal_loss+
               sibling+
               drought+
               highGrpSize+
               lowMatSCI+
               lowMatRank+
               ordrank+
               (1|sname)+
               (1|collect_grp)+
               (1|hydroyear),
             data=PredLife_M.B)
summary(model1)
#check 8Nov21

#for AlphaOrNo check 3Jun23
PredLife_M.B2<-PredLife_M.B %>% 
  mutate(AlphaOrNo=if_else(ordrank==1,"yes","no"))
model2<-lmer(age_mb~age_chrono+
               avg_month_maxtemp+
               sum_month_rain+
               season+
               maternal_loss+
               sibling+
               drought+
               highGrpSize+
               lowMatSCI+
               lowMatRank+ordrank+
               AlphaOrNo+
               (1|sname)+
               (1|collect_grp)+
               (1|hydroyear),
             data=PredLife_M.B2)
summary(model2)


PredLife_M_toprank<-PredLife_M.B %>% filter(ordrank<=median(ordrank,na.rm=T))
PredLife_M_bottomrank<-PredLife_M.B %>% filter(ordrank>median(ordrank,na.rm=T))
model3<-lmer(age_mb~age_chrono+
               avg_month_maxtemp+
               sum_month_rain+
               season+
               maternal_loss+
               sibling+
               drought+
               highGrpSize+
               lowMatSCI+
               lowMatRank+
               ordrank+
               (1|sname)+
               (1|collect_grp)+
               (1|hydroyear),
             data=PredLife_M_toprank)
summary(model3)
model4<-lmer(age_mb~age_chrono+
               avg_month_maxtemp+
               sum_month_rain+
               season+
               maternal_loss+
               sibling+
               drought+
               highGrpSize+
               lowMatSCI+
               lowMatRank+
               ordrank+
               (1|sname)+
               (1|collect_grp)+
               (1|hydroyear),
             data=PredLife_M_bottomrank)
summary(model4)


toyrank<-PredLife_M.B %>% 
  select(sname, collect_grp,hydroyear,age_mb, delta,age_chrono, avg_month_maxtemp,sum_month_rain,season, maternal_loss,sibling,drought,highGrpSize,lowMatSCI,lowMatRank,ordrank) %>% 
  mutate(age_chrono=median(age_chrono,na.rm=T),
         avg_month_maxtemp=median(avg_month_maxtemp,na.rm=T),
         sum_month_rain=median(sum_month_rain,na.rm=T),
         season="wet",
         maternal_loss=F,
         sibling=F,
         drought=F,
         highGrpSize=F,
         lowMatSCI=F,
         lowMatRank=F) %>% 
  mutate(ordrank=as.numeric(ordrank)) %>% 
  mutate(ordrank_toy=case_when(ordrank==1~1,
                               ordrank>=40~50,
                               ordrank>1&ordrank<40~median(ordrank,na.rm=T))) %>% 
  mutate(ordrank=as.integer(ordrank),
         ordrank_toy=as.integer(ordrank_toy))
# Predicts the future values
toyrank$rankpredict<-predict(model1, newdata = toyrank,allow.new.levels=T)

median(toyrank[toyrank$ordrank_toy==1,]$rankpredict,na.rm=T) #highrank, 1.332417

median(toyrank[toyrank$ordrank_toy==11,]$rankpredict,na.rm=T) #medrank, 1.059418

median(toyrank[toyrank$ordrank_toy==50,]$rankpredict,na.rm=T) #lowrank, 0.4263003
```

##### Cumulative Adversity {.tabset}
```{r}
PredLife_M.B<-PredLife.B %>% filter(sex=="M")%>% 
  mutate(cumulative_adversity=case_when(cumulative_adversity==0~0,
                                        cumulative_adversity==1~1,
                                        cumulative_adversity==2~2,
                                        cumulative_adversity>=3~3))
model<-lmer(delta~age_chrono+
               avg_month_maxtemp+
               sum_month_rain+
               season+
               cumulative_adversity+
               ordrank+
               (1|sname)+
               (1|collect_grp)+
               (1|hydroyear),
             data=PredLife_M.B)
summary(model)
#check 3Jun23
```

#### Females {.tabset}
##### Sources of Adversity {.tabset}
```{r}
PredLife_F.B<-PredLife.B %>% filter(sex=="F")
model1<-lmer(delta~age_chrono+
               avg_month_maxtemp+
               sum_month_rain+
               season+
               maternal_loss+
               sibling+
               drought+
               highGrpSize+
               lowMatSCI+
               lowMatRank+
               proprank+
               adf_collect+
               (1|sname)+
               (1|collect_grp)+
               (1|hydroyear),
             data=PredLife_F.B) 
summary(model1)
#check 3Jun23

PredLife_F_toprank<-PredLife_F.B %>% filter(proprank>=median(proprank,na.rm=T))
PredLife_F_bottomrank<-PredLife_F.B %>% filter(proprank<median(proprank,na.rm=T))
model2<-lmer(age_mb~age_chrono+
               avg_month_maxtemp+
               sum_month_rain+
               season+
               maternal_loss+
               sibling+
               drought+
               highGrpSize+
               lowMatSCI+
               lowMatRank+
               proprank+
               adf_collect+
               (1|sname)+
               (1|collect_grp)+
               (1|hydroyear),
             data=PredLife_F_toprank)
summary(model2)
model3<-lmer(age_mb~age_chrono+
               avg_month_maxtemp+
               sum_month_rain+
               season+
               maternal_loss+
               sibling+
               drought+
               highGrpSize+
               lowMatSCI+
               lowMatRank+
               proprank+
               adf_collect+
               (1|sname)+
               (1|collect_grp)+
               (1|hydroyear),
             data=PredLife_F_bottomrank)
summary(model3)


# Creating a data frame
toyrank<-PredLife_F.B %>% 
  select(sname, collect_grp,hydroyear,age_mb, delta, age_chrono, avg_month_maxtemp, sum_month_rain, season, maternal_loss, sibling, drought, highGrpSize, lowMatSCI, lowMatRank, proprank, adf_collect) %>% 
  mutate(age_chrono=median(age_chrono,na.rm=T),
         avg_month_maxtemp=median(avg_month_maxtemp,na.rm=T),
         sum_month_rain=median(sum_month_rain,na.rm=T),
         season="wet",
         maternal_loss=F,
         sibling=F,
         drought=F,
         highGrpSize=F,
         lowMatSCI=F,
         lowMatRank=F) %>% 
  mutate(proprank=case_when(proprank==1~1,
                            proprank<=.1~.1,
                            proprank<1&proprank>.1~median(proprank,na.rm=T)))
# Predicts the future values
toyrank$rankpredict<-predict(model1, newdata = toyrank,allow.new.levels=T)

median(toyrank[toyrank$proprank==1,]$rankpredict,na.rm=T) #highrank, 0.1641621

median(toyrank[toyrank$proprank==.6471,]$rankpredict,na.rm=T) #medrank, -0.2813401

median(toyrank[toyrank$proprank==.1,]$rankpredict,na.rm=T) #lowrank, -1.24212

```

##### Cumulative Adversity {.tabset}
```{r}
PredLife_F.B<-PredLife.B %>% filter(sex=="F")%>% 
  mutate(cumulative_adversity=case_when(cumulative_adversity==0~0,
                                        cumulative_adversity==1~1,
                                        cumulative_adversity==2~2,
                                        cumulative_adversity>=3~3))
model<-lmer(delta~age_chrono+
               avg_month_maxtemp+
               sum_month_rain+
               season+
               cumulative_adversity+
               proprank+
               adf_collect+
               (1|sname)+
               (1|collect_grp)+
               (1|hydroyear),
             data=PredLife_F.B) 
summary(model)
#check 3Jun23
```

# Slope over life 
#### Males {.tabset .tabset-pills}
##### Sources of Adversity {.tabset}
```{r}
model<-(lm(slope_covlife_1~
               maternal_loss+
               sibling+
               drought+
               highGrpSize+
               lowMatSCI+
               lowMatRank+
             avg_ordrank,
             data=Slope_M_life))
summary(model)
#check 8Nov21 
```

##### Cumulative Adversity {.tabset}
```{r}
Slope_M_life<-Slope_M_life %>% 
  mutate(cumulative_adversity=case_when(cumulative_adversity==0~0,
                                        cumulative_adversity==1~1,
                                        cumulative_adversity==2~2,
                                        cumulative_adversity>=3~3))
length(Slope_M_life$sname)
model<-(lm(slope_covlife_1~
             cumulative_adversity+
             avg_ordrank,
             data=Slope_M_life))
summary(model)
#check 8Nov21 
```

#### Females {.tabset .tabset-pills}
##### Sources of Adversity {.tabset}
```{r}
model<-(lm(slope_covlife_1~
               maternal_loss+
               sibling+
               drought+
               highGrpSize+
               lowMatSCI+
               lowMatRank+
             avg_proprank+avg_adf_collect,
             data=Slope_F_life))
summary(model); AIC(model)
#check 8Nov21 

cor(Slope_F_life$intercept_1,Slope_F_life$slope_covlife_1) #-0.8838758

```

##### Cumulative Adversity {.tabset}
```{r}
Slope_F_life<-Slope_F_life %>% 
  mutate(cumulative_adversity=case_when(cumulative_adversity==0~0,
                                        cumulative_adversity==1~1,
                                        cumulative_adversity==2~2,
                                        cumulative_adversity>=3~3))

model<-(lm(slope_covlife_1~
             cumulative_adversity+
             avg_proprank+avg_adf_collect,
             data=Slope_F_life))
summary(model); AIC(model)
#check 8Nov21 
```
